MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-08-19 main/MSstatsPTM_Main.tex
First final

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance ($ANOVA$)\cite{girden:1992}, or $Limma$\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications. The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the unmodified protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified proteins, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods. In particular, accounting for the confounding from unmodified protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

{\bf Simulation design:} The simulation represented an idealistic case. 24 synthetic label-free datasets were generated with different experimental designs and different biological variation. In each dataset, 1,000 proteins had 10 unmodified features per protein. Each of the 1,000 proteins had one PTM. Each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All the differential PTMs were generated with an expected log base 2 fold change of 0.75 between conditions.

@@

{\bf Simulation design:} The data were simulated as above, while providing a more realistic representation of the experiments. The feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. Specifically, PTMs were simulated with 2 modified peptide features, and unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.

@@

\medskip \noindent {\bf Data acquisition:} This experiment did not include a separate global profiling run to measure unmodified peptides. In addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate is represented by a single LC-MS/MS run.  
\figref{data-structure} schematically illustrates this data structure for one protein and one PTM site, $I=2$ and $J=2$.
For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
The number of modified and unmodified features typically varies across proteins. Some log-intensities can be outliers, and some spectral features can be missing. 
The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified protein in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in a separate enrichment and global proteome profiling run.

@@

Analysis of Variance ($ANOVA$)~\cite{kutner_etal_04a} is the simplest statistical model for summarized modified features in each biological replicate. The summarization often consists of averaging (or taking the median or other robust summary) of the log$_2$ intensities of the modified features in each replicate, e.g. $\hat{y}^*_{ij} = \sum_{k=1}^K{y_{ijk}^{\ast}}/K$. Alternatively, summarization sums the intensities of the modified features on the original scale, and then takes the log
\begin{eqnarray}
\hat{y}^*_{ij} =  \log_2 \left(\sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}\right)
\label{eq:anova_summ}
\end{eqnarray}
The basic $ANOVA$ model  is then
\begin{eqnarray}
\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{PTM}$ and its standard error. The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

@@

The estimated summaries of the modified features are used as the input to models  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$. From these summaries, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the sources of variation in the modified features that are unexplained by the model are independent from the corresponding sources of variation in the modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the unmodified and modified model fits
\begin{eqnarray}
\widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as 
\begin{eqnarray}
 \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}}
\label{eq:se_calc}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each individual feature type, reflecting the combined uncertainty in the two estimates. 
Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
\begin{eqnarray}
df \left(  \widehat{SE}(\hat{\Delta}_{adj}) \right)=  \left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

@@

The statistical framework in $MSstatsPTM$ enables sample size calculation for future experiments studying changes in PTM. 
The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively. Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow. Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\Delta_{adj})$:

\begin{equation}
SE(\Delta_{adj})^2 = \left[ \frac{2}{J} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{z_{1-\beta} + z_{1-\alpha /2}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}
and $z_{1-\beta}$ and $z_{1-\alpha /2}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Standard Normal distribution.  Solving for $J$, the number of biological replicates per condition is
\begin{equation}
J \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
\end{equation}
The numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation. Therefore, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation. Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec. \ref{supp-sec:complex_design}}. Examples of power analyses are in {\bf Supplementary Sec. \ref{supp-sec:sample_size_analysis}}.

@@

The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (Step 4 of~\figref{msstatsptm_design}). Summarization is performed separately for the PTM and the unmodified protein. When summarizing the PTM, modified peptide features that span the same modification site are summarized together. Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately. The unmodified protein summarization is performed as discussed above for $MSstats$. When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable results. The outputs of this step are the run-level summaries for the modified $\hat{y}_{ij}^{\ast}$, and unmodified $\hat{y}_{ij}$, features.

@@

After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above. Modification sites which lack corresponding global profiling information cannot be adjusted for changes in protein abundance. In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide. The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.

@@

We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$, both before and after adjusting for changes in the unmodified protein. Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was therefore excluded from the comparisons.

$MSstatsPTM$ before adjusting for changes in the unmodified protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level as described in {\bf Supplementary Table \ref{supp-fig:statistical_inference}}, as appropriate for the experimental design. $MSstatsPTM$ with the adjustment, described in \secref{adjustment_summary}, was used without imputing missing values, and without Empirical Bayes moderation.

$Unadjusted$ $ANOVA$, i.e. $ANOVA$ before adjusting for changes in the unmodified protein, was as in \eqref{ttest_model}. $Adjusted$ $ANOVA$, i.e. $ANOVA$ with the adjustment was modeled as in \eqref{ttest_ratio_model}. 
Finally, $unadjusted$ $Limma$ used the same model formula in \eqref{ttest_model}, while including a moderated variance estimation. $Adjusted$ $Limma$ was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.  
All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs. The true positives were defined as PTMs with changes distinct from the overall changes in unmodified protein abundance. The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant. Additional summaries were performed including accuracy, recall, and positive predictive value (PPV)/empirical False Discovery Rate (eFDR) as described in \eqref{summary_stats}.

@@

\figref{sim1_fdr} summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $adjusted$ $ANOVA$, $adjusted$ $Limma$ and base $MSstats$, $unadjusted$ $ANOVA$ and $unadjusted$ $Limma$ methods. The simulation mimicked a ``clean" label-free group comparison experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers. All the analyses were performed to control the eFDR at at most 5\%.  Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives. The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level.

@@

\figref{sim2_acc} summarizes the overall accuracy of methods adjusting for changes in the unmodified protein for Computer Simulation 2. The simulation mimicked a more realistic label-free group comparison experiment, including low counts of modified features and missing values. We evaluated the impact of low versus high noise, of the number of biological replicates, and of the number of conditions. For low noise, $MSstatsPTM$ outperformed the existing methods across all conditions and number of replicates, with near 100\% accuracy when the replicates were high. As the noise increased the accuracy of all the methods decreased, however $MStatsPTM$ still outperformed the existing methods. The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish. 
The ratio-based summarization approach used by $Limma$ and $ANOVA$ requires measurements for both the PTM and unmodified protein. $MSstatsPTM$ in contrast can still leverage the information in the PTM or unmodified protein if one is missing. On average over all simulations, 3.94\% of the total unmodified protein data used by $MSstatsPTM$ was thrown out by $Limma$ and $ANOVA$ due to missing PTM data. No PTM data was thrown out due to missing protein data.

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true log$_{2}$FC & Adj. true log$_{2}$FC \\ %\hline \hline
%mix2 vs mix1  & -1 & -1 \\ \hline
%mix3 vs mix1  & 0 & 1 \\ \hline
%mix4 vs mix1  & -1 & 0 \\ \hline
%mix3 vs mix2  & 1 & 2 \\ \hline
%mix4 vs mix2  & 0 & 1 \\ \hline
%mix4 vs mix3  & -1 & -1 \\ \hline
%\end{tabular}
%\end{table}

>>>>

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance ($ANOVA$)\cite{girden:1992}, or $Limma$\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide summaries, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications. The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the unmodified protein-level summaries are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified portion of the protein sequence, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods. In particular, accounting for the confounding from unmodified protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

{\bf Simulation design:} The simulation represented an idealistic case. 24 synthetic label-free datasets were generated with different experimental designs and different biological variation. In each dataset, 1,000 proteins had 10 unmodified features per protein. Each of the 1,000 proteins had one PTM. Each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified portion of the protein. Finally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified portion of the protein. All the differential PTMs were generated with an expected log base 2 fold change of 0.75 between conditions.

@@

{\bf Simulation design:} The data were simulated as above, while providing a more realistic representation of the experiments. The feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. Specifically, PTMs were simulated with 2 modified peptide features, and unmodified portions of the protein were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.

@@

\medskip \noindent {\bf Data acquisition:} This experiment did not include a separate global profiling run to quantify unmodified peptides. In addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate is represented by a single LC-MS/MS run.  
\figref{data-structure} schematically illustrates this data structure for one protein and one PTM site, $I=2$ and $J=2$.
For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
The number of modified and unmodified features typically varies across proteins. Some log-intensities can be outliers, and some spectral features can be missing. 
The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified portion of the protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified portion of the protein in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in a separate enrichment and global proteome profiling run.

@@

Analysis of Variance ($ANOVA$)~\cite{kutner_etal_04a} is the simplest statistical model for summarized modified features in each biological replicate. The summarization often consists of averaging (or taking the median or other robust summary) of the log$_2$ intensities of the modified features in each replicate, e.g. $\hat{y}^*_{ij} = \sum_{k=1}^K{y_{ijk}^{\ast}}/K$. Alternatively, summarization sums the intensities of the modified features on the original scale, and then takes the log
\begin{eqnarray}
\hat{y}^*_{ij} =  \log_2 \left(\sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}\right)
\label{eq:anova_summ}
\end{eqnarray}
The basic $ANOVA$ model  is then
\begin{eqnarray}
\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{PTM}$ and its standard error. The estimates are used to test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified portion of the protein.

@@

The estimated summaries of the modified features are used as the input to models  such as in~\eqref{msstats_model} or~\eqref{msstatstmt_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$. From these summaries, the proposed approach estimates the adjusted difference $\hat{\Delta}_{adj}$ in~\eqref{null_hyp}
\begin{eqnarray}
\hat{\Delta}_{adj} = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i}) - (\hat{\mu}_{i^{\prime}}^{\ast}-\hat{\mu}_{i^{\prime}}) = (\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}) - (\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}) = \hat{\Delta}_{PTM} - \hat{\Delta}_{protein} 
\label{eq:Delta_hat_adj}
\end{eqnarray}
Assuming that the sources of variation in the summaries of modified features that are unexplained by the model are independent from the sources of variation in the summaries of modified features, the standard error $\widehat{SE}(\hat{\Delta}_{adj})$ is obtained by combining the standard errors from the two model fits
\begin{eqnarray}
\widehat{SE}(\hat{\Delta}_{adj}) = \sqrt{ \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2}} 
\label{eq:prop_se}
\end{eqnarray}
For example, in the simple case of \figref{data-structure} with $J=2$ conditions, where $\hat{\sigma}^2_{PTM}$ and $\hat{\sigma}^2_{protein}$ are respectively  the estimates of the error variance for the PTM and protein model described in \eqref{msstats_model}, the standard error is calculated as 
\begin{eqnarray}
 \widehat{SE}(\hat{\Delta}_{adj}) &=& \sqrt{\widehat{SE}(\hat{\Delta}_{PTM})^2 + \widehat{SE}(\hat{\Delta}_{protein})^2} = \sqrt{\frac{1}{J}\hat{\sigma}^2_{PTM} + \frac{1}{J}\hat{\sigma}^2_{protein}}
\label{eq:se_calc}
\end{eqnarray}
The estimated standard error is larger than the standard errors associated with each individual feature type, reflecting the combined uncertainty in the two estimates. 
Finally, the degrees of freedom associated with~\eqref{prop_se} are obtained via the Satterthwaite approximation~\cite{kutner_etal_04a, satterthwaite:1946}
\begin{eqnarray}
df \left(  \widehat{SE}(\hat{\Delta}_{adj}) \right)=  \left( \widehat{SE}(\hat{\Delta}_{PTM})^{2} + \widehat{SE}(\hat{\Delta}_{protein})^{2} \right)^2 \bigg/
\left( \frac{\widehat{SE}(\hat{\Delta}_{PTM})^{4}}{\mathrm{df}(\widehat{SE}(\hat{\Delta}_{PTM}))} + \frac{\widehat{SE}(\hat{\Delta}_{protein})^{4}}{ \mathrm{df}(\widehat{SE}(\hat{\Delta}_{protein}))} \right)
\label{eq:prop_df}
\end{eqnarray}
To test the null hypothesis in~\eqref{null_hyp}, the test statistic $\hat{\Delta}_{adj}/\widehat{SE}(\Delta)_{adj}$ is compared with the Student distribution with the degrees of freedom in~\eqref{prop_df}. The p-values of the comparison are adjusted for multiple testing using the approach by Benjamin and Hochberg~\cite{Benjamini:1995}.

@@

The statistical framework in $MSstatsPTM$ enables sample size calculation for future experiments studying changes in PTM. 
The procedure has been described in general in~\cite{kutner_etal_04a}, and for protein significance analysis specifically in~\cite{oberg_vitek_09a}. 
It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the pairwise comparisons of conditions for the modified and unmodified protein summaries, respectively. Typically, these variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow. \todo{The estimation of the variance components for pairwise comparisons of conditions depends on the experimental design, and is detailed in Supply Figure S1}

Given the above quantities, and assuming a balanced design that allocates the same number of replicates $J^{\prime}$ to both modified and unmodified profiling runs, the minimal number of replicates $J^{\prime}$ for each of $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\Delta_{adj})$:

\begin{equation}
SE(\Delta_{adj})^2 = \left[ \frac{2}{J^{\prime}} \left( \sigma_{PTM}^{2} + \sigma_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{z_{1-\beta} + z_{1-\alpha /2}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}
and $z_{1-\beta}$ and $z_{1-\alpha /2}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Standard Normal distribution.  Solving for $J^{\prime}$, the number of biological replicates per condition is
\begin{equation}
J^{\prime} \geq \frac{(2\sigma_{PTM}^{2} + 2\sigma_{protein}^{2})(z_{1-\beta} + z_{1-\alpha /2})^2}{\Delta_{adj}^2}
\end{equation}
The numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation. Therefore, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.

\todo{Here we calculate sample size, but in the results we calculate power. Could you briefly comment on how we can solve for the power given the sample size (say that this is using the same equation, but solving for $q$ instead.) In fact we can fix all value but one and solve for the remainder}

\todo{I rearranged the supplementary sections. Could you check throughout the manuscript that all the supplementary sections are appropriately referenced?}

@@

The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (Step 4 of~\figref{msstatsptm_design}). Summarization is performed separately for the PTM features, and for the features representing the unmodified portion of the protein. When summarizing the PTM, modified peptide features that span the same modification site are summarized together. Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately. The unmodified protein summarization is performed as discussed above for $MSstats$. When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable results. The outputs of this step are the run-level summaries for the modified $\hat{y}_{ij}^{\ast}$, and unmodified $\hat{y}_{ij}$, features.

@@

After modeling, the model for the modified peptides is adjusted for changes in the abunance of the unmodified portion of the protein, using the methods described above. Modification sites which lack corresponding global profiling information cannot be adjusted for changes in protein abundance. In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide. The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.

@@

We compared the performance of $MSstatsPTM$ to that of $Limma$ and $ANOVA$, both before and after adjusting for changes in the unmodified portion of the protein. Since $Isobar$-$PTM$ is only applicable to experiments with TMT labeling it could not be applied to the datasets with known ground truth in this manuscript, and was therefore excluded from the comparisons.

$MSstatsPTM$ before adjusting for changes in the unmodified portion of the protein corresponds to base $MSstats$ or $MSstatsTMT$, modeled on the peptide level as described in {\bf Supplementary Table \ref{supp-fig:statistical_inference}}, as appropriate for the experimental design. $MSstatsPTM$ with the adjustment described in \secref{adjustment_summary} was used without imputing missing values, and without Empirical Bayes moderation.

$Unadjusted$ $ANOVA$, i.e. $ANOVA$ before adjusting for changes in the unmodified portion of the protein, was as in \eqref{ttest_model}. $Adjusted$ $ANOVA$, i.e. $ANOVA$ with the adjustment was modeled as in \eqref{ttest_ratio_model}. 
Finally, $unadjusted$ $Limma$ used the same model formula in \eqref{ttest_model}, while including a moderated variance estimation. $Adjusted$ $Limma$ was modeled as in \eqref{ttest_ratio_model} including moderated variance estimation.  
All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs. The true positives were defined as PTMs with changes distinct from the overall changes in abundance of the unmodified portion of the protein. The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant. Additional summaries were performed including accuracy, recall, and positive predictive value (PPV)/empirical False Discovery Rate (eFDR) as described in \eqref{summary_stats}.

@@

\figref{computer_sim}(a) summarizes the eFDR reported on Computer Simulation 1 dataset by $MSstatsPTM$, $adjusted$ $ANOVA$, $adjusted$ $Limma$ and base $MSstats$, $unadjusted$ $ANOVA$ and $unadjusted$ $Limma$ methods. The simulation mimicked a ``clean" label-free group comparison experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers. All the analyses were performed to control the eFDR at at most 5\%.  Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives. The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level.

@@

\figref{computer_sim}(b) summarizes the overall accuracy of methods adjusting for changes in abundance of the unmodified portion of the proteins in Computer Simulation 2. The simulation mimicked a more realistic label-free group comparison experiment, including low counts of modified features and missing values. We evaluated the impact of low versus high noise, of the number of biological replicates, and of the number of conditions. For low noise, $MSstatsPTM$ outperformed the existing methods across all conditions and number of replicates, with near 100\% accuracy when the replicates were high. As the noise increased the accuracy of all the methods decreased, however $MStatsPTM$ still outperformed the existing methods. The difference was primarily due to two reasons. First, the ratio-based summarization approach used by $Limma$ and $ANOVA$ requires measurements for both the PTM and unmodified protein. In contrast, $MSstatsPTM$ can leverage the information in the PTM or unmodified portion of the protein if one of the two is missing. On average over all simulations, 3.94\% of the total unmodified protein data \todo{what do you mean by 'data'? Runs? Combination of features and runs?} used by $MSstatsPTM$ was discarded by $Limma$ and $ANOVA$ due to missing PTM data. No PTM data was discarded due to missing protein data. Second, the robust Turkey Median Polish summarization in $MSstatsPTM$ was more resistant to outliers.

@@
